\emph{If I have seen further, it is by standing on the shoulders of giants.}
\\
\\
\emph{Isaac Newton}
\\

I must begin by thanking my doctoral advisor, Gill Bejerano. During my time in the lab, he has taught me the importance of asking the \emph{right} questions. Reflecting on this lesson, it seems to be only a slight variation on one of life's most important questions: What should we think about? Thank you.

Aaron Wenger is the second giant on whose shoulders I stand. Aaron was my student mentor in the lab, and his generosity, meticulousness, and keen sense for finding the right path to answer questions profoundly shaped me as a researcher. Thank you.

I would like to thank the other members of the Bejerano lab: Saatvik Agarwal, Mark Berger, Johannes Birgmeier, Gray Camp, Charles Chen, Jenny Chen, Sandeep Chinchali, Tisha Chung, Shoa Clarke, Andrew Doxey, Harendra Guturu, Whitney Heavner, Michael Hiller, Karthik Jagadeesh, Robin Jia, Amir Markovitz, Cory McLean, Ravi Parikh, Tomas Rube, Bruce Schaar, Sushant Shankar, Sanjay Siddhanti, Yatish Turakhia, Geetu Tuteja, and Lindsay Willmore. In particular, Harendra Guturu and Geetu Tuteja played an outsized role in helping me think about different problems. 

I would also like to thank my thesis reading committee, Susan McConnell, Philippe Mourrain, and David Dill. Sue taught me the importance of writing - and thus thinking - clearly. My first collaboration with Philippe resulted in all of my predictions failing to validate. His faith and optimism in computational methods to complement biological experiments resulted in some of my most beautiful doctoral work.

I thank the National Science Foundation and the Office of the Vice Provost for Graduate Education for financial support. The freedom afforded by this assistance allowed me to pursue questions I would not have otherwise asked.

Finally I would like to thank my friends and family. If the above colleagues are the giants on whose shoulders I stand, my friends and family are those who have helped me climb there. I especially thank my parents, Lindsay and Mary, and my brother, Alex. I will forever be indebted to my many great teachers, but I count my parents the first and greatest among them.

The results presented in \chapref{chap:mer130} and \chapref{chap:autism} were collaborative efforts that derive from previously published work:

\begin{enumerate}
\item James H. Notwell, Tisha Chung, Whitney E. Heavner, and Gill Bejerano. A family of transposable elements co-opted into developmental enhancers in the mouse neocortex. \emph{Nature Communications}, \textbf{6}, 6444 (2015).

\item James H. Notwell, Whitney E. Heavner, Siavash Fazel Darbandi, Sol Katzman, William L. McKenna, Christian F. Ortiz-Londono, David Tastad, Matthew J. Eckler, John L.R. Rubenstein, Susan K. McConnell, Bin Chen, and Gill Bejerano. TBR1 regulates autism risk genes in the developing neocortex. \emph{Genome Research}, \textbf{26}, 1013-1022. (2016).
\end{enumerate}

In the \chapref{chap:mer130} work, Tisha Chung and Whitney Heavner performed all transfection experiments. I designed the study together with Gill Bejerano, performed the computational analyses, and wrote the manuscript with Whitney Heavner and Gill Bejerano.

In the \chapref{chap:autism} work, Bin Chen and members of her lab, William McKenna, Christian Ortiz-Londono, David Tastad, and Matthew Eckler, collected tissue samples and performed ChIP-seq. Whitney Heavner and Siavash Fazel Darbandi performed the \emph{in situ} experiments. I designed the study together with Bin Chen and Gill Bejerano, performed the computational analyses with advice from Sol Katzman, and wrote the manuscript with Whitney Heavner, Siavash Fazel Darbandi, John Rubenstein, Susan McConnell, Bin Chen, and Gill Bejerano.

\chapref{chap:zfishSnps} was also a collaborative effort and has been submitted for publication. Romain Madelaine performed all the \emph{in vivo} experiments with technical support from Gemini Skariah and Caroline Halluin. I designed the computational study with Gill Bejerano, performed the computational analyses with assistance from Charles Chen, and wrote the manuscript with Romain Madelaine, Gill Bejerano, and Philippe Mourrain.

