I am very grateful to my doctoral advisor, Prof. Gill Bejerano, who has been a wonderful mentor and friend.  Gill taught
me the technical, research, and managerial skills to succeed in graduate research.  Additionally, he imparted on
me his love of genomics, and he provided a nurturing and pleasant working environment.  I appreciate the opportunities and
education he has given me, and the work of this dissertation is as much his as mine.
%

I would like to thank many past and present members of the Bejerano Lab for their contributions: Saatvik Agarwal, Cory Barr,
Dave Bristor, Jenny Chen, Tisha Chung, Shoa Clarke, Andrew Doxey, Harendra Guturu, Michael Hiller, Cory McLean, Karen Moreira,
Jim Notwell, Zaneta Odrowaz, Ravi Parikh, Hrysoula Papadakis, Tomas Rube, Sushant Shankar, Bruce Schaar, and Geetu Tuteja.
%

I have worked especially closely with Shoa Clarke, Harendra Guturu, and Cory McLean, and I am thankful for their ideas
and contributions.  Shoa provided helpful feedback on all of the tools that underlie this dissertation, and he inspired many of
the ideas presented herein.  Harenda assisted in development of my tools, and I applied many utilities that he developed
independently.  He also contributed ideas for the analysis presented in this dissertation.  Cory developed the GREAT
approach on which I relied heavily, and he provided technical education and assistance as well as ideas for analysis.
It was a pleasure to work with all three, and I feel lucky to have had the opportunity.
%

I would like to thank Geetu Tuteja and Bruce Schaar, who led the experimental efforts in the Bejerano Lab during my time.
Their experiments provided valuable data for analysis, and their insights and biological perspective assisted me in viewing
my work from unique angles.  I am also thankful to Michael Hiller, who contributed numerous ideas and provided a novel
perspective.
%

I appreciate the time and ideas offered by Prof. David Kingsley and Prof. William Talbot.  Both provided useful comments
and ideas in person and in seminars and assisted me with interpreting experimental data.
%

Lastly, I would like to thank my friends and family for support.  Specifically, I am thankful to my mom Sue, my dad
Allen, my sisters Rachel and Hannah, my brother Jacob, and my girlfriend Susie James.
%

The works presented in \chapref{chap:prism} and \chapref{chap:neocortex} were collaborative efforts and have
been prepared as manuscripts for journal publication.
%

In \chapref{chap:prism}, 
Shoa Clarke contributed valuable ideas and comments, performed the literature search to
connect PRISM predictions to known literature, and wrote the corresponding sections of the manuscript.  Jenny Chen
gathered and annotated the library of transcription factor binding motifs used in the PRISM screen.  Cory McLean provided
useful comments and helped develop the PRISM web portal.  Gill Bejerano 
oversaw the research and co-wrote the manuscript.  I developed the excess conservation and PRISM approaches, implemented them
as a set of computer programs, designed and performed most of the analysis, built most of the PRISM web resource, and co-wrote the manuscript.
%

In \chapref{chap:neocortex},
Bruce Schaar performed most of the experimental work with assistance from Tisha Chung and Geetu Tuteja.  Bruce also
designed the experiments with me, Shoa Clarke, and Gill Bejerano.  Shoa Clarke performed an initial motif enrichment
analysis using a tool built by Harendra Guturu.  I designed and performed most of the computational analysis and co-wrote the
manuscript with Gill Bejerano.  With Shoa Clarke, I will be preparing a future version of the work for publication.

