Much of the functional sequence in the human genome is devoted to
regulatory elements, which control when during development and where in
the body genes are active.  \textit{Cis}-regulatory elements have been shown to
contribute to human evolution and disease, but the function of most regulatory elements in
the human genome is unannotated, and the grammatical rules for how regulatory
function is encoded in DNA sequence are unknown.  This dissertation
describes the application of a suite of novel computational genomics tools
in 1) a genome-wide screen to broadly annotate the
binding site composition and biological function of regulatory elements in the human and
mouse genomes, and 2) a study of development of the mouse
neocortex, a mammalian-specific brain region that underlies higher cognitive functions.
%

The genome-wide screen, termed PRISM for ``Predicting Regulatory Information
from Single Motifs'', combines accurate transcription factor
binding site prediction using a novel ``excess conservation score'' with
GREAT, a tool for functional enrichment analysis of regulatory regions.  The
excess conservation score identifies binding sites that are conserved more
strongly than surrounding nucleotides, a signature which suggests specific
evolutionary constraint on the binding site.  Binding site prediction with
the excess conservation score outperforms the previous state-of-the-art for
44 of 47 examined transcription factors.  The genome-wide PRISM screen identifies a biological function
for 109,748 mostly distal binding sites in regulatory elements associated with 7,692
different target genes at an overall false discovery rate of 15\%.  Many of
the predictions are supported by prior literature, but the majority of predictions
are novel and thus provide hypotheses for further experimental study.  A web
portal -- http://PRISM.stanford.edu -- presents the data to biomedical researchers.
%

The case study of neocortex development combines computational analysis of
regulatory elements with high-throughput sequencing techniques to uncover
gene interaction networks in a rapid, non-targeted manner that is not
possible with classical genetics methods.  ChIP-seq for the enhancer-associated
co-activator protein p300 in dissected neocortex from E14.5 mouse embryos
identifies 6,629 candidate enhancer regions.  The candidate enhancers are
enriched near genes expressed in E14.5 neocortex and important to neocortex development.
Eight of ten candidates tested in a mouse transgenic enhancer assay drive activity in
specific patterns within the neocortex.  Surprisingly, 23\% of the 6,629 candidates
are evolutionarily conserved beyond mammals and thus pre-date the mammalian innovation of
the neocortex.  Motif analysis identifies \textit{Neurod}, \textit{Lhx/Lmx}, \textit{Nfi},
\textit{Rfx}, and \textit{Ctcf}
motifs enriched within the candidate enhancers, and the motifs are connected to specific
functions and diseases.  I propose an approach to utilize such a combination of
computational binding site analysis and functional genomics enhancer assays
to identify tissue-specific gene regulatory networks.  I demonstrate the approach by
predicting targets of \textit{Rbpj}.
%

These studies extend our knowledge of regulatory elements in the human genome, provide a suite
of novel methods and tools, and illustrate how
characterization of regulatory elements offers a novel perspective to genetic and genomic studies.
